---
title: "Wind Summer Steelhead Mark-Resight and \"jumper\" ratio estimates from August
  to September"
author: Thomas Buehrens (tbuehrens@dfw.wa.gov) & Charlie Cochran (patrick.cochran@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***

```{r setup, include=FALSE,echo = TRUE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Estimates of Wind River wild summer steelhead runsize above Shipherd Falls are made by tagging the subset of summer steelhead that use the fish ladder (while others jump the falls) as they pass upstream. Riverscape snorkel surveys are conducted in August and September to enumerate tagged and un-tagged steelhead upstream of the falls, and enable abundance estimation via a Lincoln-Petersen estimate.These estimates are incomplete because steelhead continue to pass the falls after the September survey. Sometime in the fall, usually in October, a combination of high flows and decreasing temperatures preclude summer steelhead from jumping the falls from that point on until the migration has ended for that spawn year in the following spring. However, from the September snorkel until the end of the jumping period, the falls trap count is not a census count and therefore must be expanded to account for steelhead that jumped the falls and were not caught in the trap.To accomplish this a jumper ratio is calculated as:

Equation. 1):

(September Snorkel Abundance - August Snorkel Abundance) / (September Snorkel Trap Count - August Snorkel Trap Count)

This "jumper ratio" is then multiplied by trap counts between the September snorkel and the end of the jumping period to account for untrapped fish.

The final abundance estimate is the sum of:
1) the September snorkel abundance
2) the jumper ratio expanded trap count between September and the end of the jumping period
3) a census count of steelhead trapped between the end of jumping and the end of the run.

The actual equations used to calculate the jumper ratio differ from equation 1 in order to account for and propagate uncertainty in abundance estimates, and noise in the trap counts due to discreteness. Full details are shown in the data and model appendices below.
```{r functions and packages,message = FALSE, warning = FALSE,results = "hide"}
#function to install or load packages
logit<-function(x){log(x/(1-x))}
  
install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg))
    install.packages(create.pkg, dependencies = TRUE,repos = "http://cran.us.r-project.org")
  sapply(pack, require, character.only = TRUE)
}

packages<-c("tidyverse","rstan","ggplot2","kableExtra","lubridate")
install_or_load_pack(packages)
```

```{r load data,message = FALSE, warning = FALSE,results = "hide"}


dat<-read_csv("data/jumper_data.csv")%>%
  dplyr::rename(
    tr = trap_count_septminusaug,
    N_aug_obs = august_estimate,
    N_sept_obs = september_estimate, 
  )%>%
  mutate(
    N_aug_sd_obs = sqrt(log(august_cv^2+1)) ,
    N_sept_sd_obs = sqrt(log(september_cv^2+1))
  )


years = min(dat$year) : max(dat$year)
years = years - min(years) + 1
yrs=length(years)
a_years = years[!is.na(dat$N_aug_sd_obs)]
a_yrs = length(a_years)
s_years = years[!is.na(dat$N_sept_sd_obs)]
s_yrs = length(s_years)
obs_years = s_years[s_years%in%a_years]
obs_yrs = length(obs_years)

stan_dat<-dat%>%
  as.list()

stan_dat$years = years
stan_dat$yrs = yrs
stan_dat$a_years = a_years
stan_dat$a_yrs = a_yrs
stan_dat$s_years = s_years
stan_dat$s_yrs = s_yrs
stan_dat$obs_years = obs_years
stan_dat$obs_yrs = obs_yrs

stan_dat<-lapply(stan_dat,function(x) x[!is.na(x)])
```

```{r fit stan model,message = FALSE, warning = FALSE,results = "hide"}
# run the MCMC 
if(!file.exists("results/jumper_results.rds")){
  start.time<-Sys.time()
  if(!file.exists("model/model.rds")){  
    model<-stan_model("model/jumper.stan")
    saveRDS(model,"model/model.rds")
  }else{
    model<-readRDS("model/model.rds")
  }
m1 <-sampling(
    object = model,
    data=stan_dat,
    #inits=0,
    cores = 4,
    chains=4, #Number of chains
    iter=4000, #number of iterations
    warmup=2000, #number of burn in
    thin=1 #thinning rate
  )
  end.time<-Sys.time();print(end.time-start.time)
  
  #save jags run, results
  saveRDS(m1,"results/jumper_results.rds")
  write.csv(summary(m1)$summary,"results/summary.csv")
}else{
  m1<-readRDS("results/jumper_results.rds")
}
```
## Results

Below are plots and tables of the "jumper ratio" and model smoothed vs. raw "observed" mark-recapture abundance estimates in August and September.
```{r results,message = FALSE, warning = FALSE,results = "show"}
results<-rstan::extract(m1)

jr<-t(apply(1/results$p_tr,2,function(x) quantile(x,c(0.025,0.25,0.5,0.75,0.975))))%>%
  as_tibble()%>%
  bind_cols(tibble(year=min(dat$year):max(dat$year)))

ggplot(jr, aes(x=year,y=`50%`))+
  geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_ribbon(aes(ymin=`25%`,ymax=`75%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_line()+
  ylab("Jumper Ratio")+
  theme_bw()+
  scale_y_log10(limits=c(1,40))+
  annotation_logticks(side="l")

  
jr%>%
  dplyr::rename(spawn_year = year)%>%
  kbl(caption = "Table 1.The ratio of wild steelhead jumping Shipherd Falls to the number caught in the trap between the August and September snorkels (aka the `Jumper Ratio')",digits =3)%>%
  kable_classic(full_width = F, html_font = "Cambria")

hier_jr<-quantile(1/results$hier_p_tr,c(0.025,0.25,0.5,0.75,0.975))%>%
  as.data.frame()%>%
  rownames_to_column()%>%
  dplyr::rename(quantile ="rowname",hier_jr=".")%>%
  as_tibble()

hier_jr%>%
  kbl(caption = "Table 2.The hierarchical jumper rate" ,digits =3)%>%
  kable_classic(full_width = F, html_font = "Cambria")

N_aug<-t(apply(results$N_aug,2,function(x) quantile(x,c(0.025,0.25,0.5,0.75,0.975))))%>%
  as_tibble()%>%
  bind_cols(tibble(year=min(dat$year):max(dat$year)))%>%
  left_join(dat%>%dplyr::select(year,N_aug_obs,august_l95,august_u95))

ggplot(N_aug, aes(x=year,y=`50%`))+
  geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_ribbon(aes(ymin=`25%`,ymax=`75%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_line()+
  geom_errorbar(aes(ymax = august_u95,ymin=august_l95),width = 0.25)+
  geom_point(aes(x=year,y=N_aug_obs))+
  ylab("Abundance")+
  theme_bw()+
  ylim(0,NA)

N_aug%>%
  dplyr::rename(spawn_year = year)%>%
  kbl(caption = "Table 3.The smoothed  & observed August mark-resight abundance of wild Wind River summer steelhead",digits =3)%>%
  kable_classic(full_width = F, html_font = "Cambria")

N_sept<-t(apply(results$N_sept,2,function(x) quantile(x,c(0.025,0.25,0.5,0.75,0.975))))%>%
  as_tibble()%>%
  bind_cols(tibble(year=min(dat$year):max(dat$year)))%>%
  left_join(dat%>%dplyr::select(year,N_sept_obs,september_l95,september_u95))

ggplot(N_sept, aes(x=year,y=`50%`))+
  geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_ribbon(aes(ymin=`25%`,ymax=`75%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_line()+
  geom_errorbar(aes(ymax = september_u95,ymin=september_l95), width = 0.25)+
  geom_point(aes(x=year,y=N_sept_obs))+
  ylab("September Abundance")+
  theme_bw()+
  ylim(0,NA)

N_sept%>%
  dplyr::rename(spawn_year = year)%>%
  kbl(caption = "Table 4.The smoothed & observed September mark-resight abundance of wild Wind River summer steelhead",digits =3)%>%
  kable_classic(full_width = F, html_font = "Cambria")

wdat<-readr::read_delim(paste0("https://waterdata.usgs.gov/nwis/monthly?site_no=14105700&agency_cd=USGS&por_14105700_113462=546282,00010,113462,1996-06,",year(Sys.Date()),"-",month(Sys.Date()),"&referred_module=sw&format=rdb"),comment = '#')%>%
  filter(agency_cd=="USGS" & month_nu == 8)%>%
  mutate(year=year_nu+1,mean_water_temp=as.numeric(mean_va))%>%
  dplyr::select(year,mean_water_temp)


p_aug<-t(apply(results$p_aug,2,function(x) quantile(x,c(0.025,0.25,0.5,0.75,0.975))))%>%
  as_tibble()%>%
  bind_cols(tibble(year=min(dat$year):max(dat$year)))%>%
  left_join(wdat)

ggplot(p_aug, aes(x=year,y=`50%`))+
  geom_ribbon(aes(ymin=`2.5%`,ymax=`97.5%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_ribbon(aes(ymin=`25%`,ymax=`75%`),fill="cadetblue",color="NA",alpha=0.5)+
  geom_line()+
  ylab("August Abundance as a proportion of September Abundance")+
  theme_bw()+
  ylim(0,1)

p_aug%>%
  dplyr::rename(spawn_year = year)%>%
  kbl(caption = "Table 5.August abundance as a proportion of September abundance of wild Wind River summer steelhead",digits =3)%>%
  kable_classic(full_width = F, html_font = "Cambria")

# ggplot(p_aug, aes(x=mean_water_temp,y=logit(`50%`)))+
#   geom_point()+
#   ylab("logit(p_aug)")+
#   geom_smooth(method='lm', formula= y~x)+
#   theme_bw()
# 
# summary(lm(logit(`50%`)~mean_water_temp,data=p_aug))
# 
#auto.arima(logit(p_aug$`50%`),xreg=as.matrix(p_aug$mean_water_temp))

```

## Appendices
Below is the model used to estimate the jumper ratio, and the formatted data as supplied to the model
```{r model and data,message = FALSE, warning = FALSE,results = "show"}
noquote(read_lines("model/jumper.stan"))

print(stan_dat)
```